The Nature of Breath: A Biomimetic Kinetic Sculpture that Breathes

### 2021-12-17 finalproject/report/ms1report.tex
phew

The question of what it means to be alive is unavoidable when considering the relationship between movement and what is natural. Movement is a key indicator of life and the ability to distinguish between the living and non-living is a deeply rooted capacity of the human brain. Psychologists and Neurologists have long been aware that ``the distinction between living and non-living things forms a natural place,'' to divide the functionalities and capacities of the mind.\cite{narine} Kinetic sculpture layers these two distinctions -- between the natural and the artificial and the living and non-living -- in an open-ended way which scientific disciplines are rarely allowed.

@@

Finally, it would be an oversight not to mention the hundreds of engineers and makers who devoted much time and though to the development of OpenSource emergency ventilators during the beginning and worst months of the pandemic.\cite{read} Very few of these devices were ever used on patients, but the collective thought and conversation surrounding the mechanics of breath cannot be discounted as useless, if only as a poignant reminder of the body's mystery and vulnerability.

@@

Storage describes the way air is stored and contained during the act of breath. This is done by the lungs and is only necessary for the transfer of oxygen to the blood. While the lungs are arguably the most critical part of the respiratory system, they are the most difficult to simulate with accuracy and have relatively little bearing on the sound of breath in this context.

@@

Modulation is the way the sound of breath is shaped as it leaves the body. This is achieved by the bronchial tree, trachea, larynx, pharynx, and mouth.\cite{sarkar} After actuation has been achieved, simulation of this system is the most relevant and interesting part of this project.

@@

Focusing on the two main areas of inquiry, the making and testing has been divided into been auditory prototypes -- exploring the effect of disembodied breath noises -- as well as mechanical prototypes -- exploring ways to achieve a physiologically accurate breathing mechanism. In addition to creating a visual tie between the biological nature of the respiratory system, I chose to closely mimic the diaphragm and trachea to prevent the noise of the mechanical components from overwhelming the noise of the breath produced. This approach relies on the material properties of latex and vinyl -- elasticity and rigidity -- to be functional.

@@

To test the effect of disembodied breath on strangers, a small speaker was placed inside of two five gallon plastic buckets. An old cellphone was pre-loaded with a three minute sound effect track of different types of breathing noises. These cycled through breath which sounded like someone sleeping to someone wheezing from exertion. In one bucket, a hole was drilled and a large diameter latex tube was inserted. This provided a way to direct the sound emanating from the speaker and added obfuscated the function of the assembly. The assembly was then placed near a crosswalk with a high level of foot traffic around 8 PM ET on the day of Halloween. This location was chosen specifically because it was easy to observe discreetly and had a wide variety of pedestrians.

@@

The objective of this second category of trials was to investigate ways of generating breathing noises mechanically. Using the systems breakdown discussed above, I attempted to create mechanisms for both modulation and actuation of breath. I first approached the diaphragm since actuation was a necessary step to fully explore modulation.

Using the five gallon buckets earlier, I cut out the center of the lid of one bucket and used it to secure a large piece of 25 mil natural latex over the opening. I then inserted the same large piece of latex tubing into a hole at the other end of the bucket. By depressing the latex air was forced out of the tube. Pinching the tube at different heights and changing the cadence of the diaphragm movement created a pretty wide range of breath-like noises.

@@

After measuring the force required to depress the lever fully -- 100 N -- I determined that creating an electronic system to activate the diaphragm would be costly and complicated given the scope of time. A few attempts at creating mechanical advantage for the lever were attempted, but none had successful results. It was very difficult to generate the required amount of force without putting a significant amount of weight on the lever itself. I ultimately decided that, due to the expense of high-powered and quiet motors, it made more sense to solve this problem after the core interaction of the piece had been refined.

@@

I embedded four sections of this tubing between two sheets of fourteen mil natural latex. I then bent this into a curved tube-like shape and enclosed the open end of the C FIG. Once this basic shape was complete I threaded each piece of tubing with monofilament and anchored the assembly between two dowels. By pulling on the monofilament is was possible to contract each ring within the tube individually. This result was highly successful and it was possible to contract the aperture to varying degrees.

@@

While this result was not integrated into the modulation prototype, it was by far the most naturalistic movement produced during the investigation. In comparison to the latex tubing used to mimic the trachea earlier, this mechanical latex assembly has much thinner walls. Overall, I believe this will produce more accurate breath noises when combined with the actuation mechanism

@@

I first created a digital breath generator in order to create a more nuanced interaction between the viewer and the breathing noises. This was done using P5.js and allowed me to change the cadence, pitch, and volume of the breathing noise. A simple infra-red distance sensor was then connected to the generator. As the distance between the viewer and the work decreased the breathing increased in intensity. The breath noise was amplified using a class D audio amplifier and played through speakers mounted on the front of the piece.

@@

The mechanical form of the work stayed the same in this trial, however I removed the five gallon bucket in favor of a clear vessel which allowed the viewer to see the diaphragm moving. I also added an abstracted bronchial tree inside the vessel to suggest lungs. In combination with the added electronics this had a much more polished effect.

@@

Initial research into the effects of disembodied breath revealed that the intensity and volume of the breathing noise has a large impact on the comfort level of the viewer. It was clear from the repetition of these trials in different settings that volume played major a role in the level of discomfort. Loud breathing noises inherently felt closer and more intimate than quieter noises. Cadence and pitch of breathing noises also appeared to have a lot of bearing on the emotion of the observer. A fast cadence of slightly higher pitch at higher volume was more noticeable and produced a higher degree of discomfort than a slower, softer breathing noise.

In trials which involved playing noises of ``sick'' breathing noises, viewer response was much more intense. Results of these auditory trials were interesting because they illustrated just how sensitive the group had become to the noise of respiratory distress. While direct comparison is not possible, I wonder if these experiments would have had the same effect before the COVID-19 pandemic.

These results were very helpful as I began to develop the digital breath generator. The ability to dynamically change the intensity and volume of breathing noises based on user proximity created a much more intuitive interaction between the viewer and the breathing noise. In prior user interviews I found that having the intensity of the breathing increase in the presence of the viewer was more natural than the intensity of breath decreasing. This proved to be true as the change in intensity of breathing piqued viewers interest, causing them to approach the work. When they were very close, the breathing reached its highest level of intensity and many viewers noted that they felt somewhat uncomfortable to be in such close proximity to the piece. In comparison, some viewers mentioned that at far away distances -- producing slow and relaxed breathing noises -- they found their own breath falling into a relaxing rhythm and were enjoyed the meditative quality produced. Overall, the auditory interaction consistently caused viewers to consider their own breath.

@@

By replacing the lever mechanism with a crank shaft -- a technique for translating circular motion into linear motion -- I hope to have fine grained control over the movement of the diaphragm and thereby the actuation of air through the work. This will then allow for further development and testing of the mechanical trachea. The modulation of breath does not end with the trachea however, and I hope to continue exploring the ways in which this sound can be manipulated. Perhaps the development of a mechanical pharynx, or larynx would follow.

@@

The primary objective of this work was to create a kinetic sculpture which appeared to be alive and breathing, thereby asking what it means for an non-living, artifical construct to breathe in a natural, life-like way. This objective was approached trough a combination of auditory and mechanical explorations. These investigations revealed that breath is a sensitive topic and can be interpreted as an indicator of many things. In addition to my intent of using breath as an indicator of life, I found that it is also a subtle and powerful indicator of emotion and health. The ongoing COVID-19 pandemic has certainly made us more attuned to the subtle changes in breath that signal respiratory distress.

The sound of breath can be used to great effect, however, these investigations revealed that the movement of breath is a much more intuitive signifier of life. The mechanical goal of this work was to isolate the components of the respiratory system which form the noise created by breathing and highlight their movement. The intensity of viewer's reactions to the mechanical system developed confirmed that the movement is just as powerful as sound in indicating life.

Each human breath on this planet has moved us one step further into the Anthropocene and closer, some would argue to the end of nature. To conceptualize the impact of human activity on this planet is practically impossible, yet we find ourselves at a juncture where this definition is at the heart of nearly every social and political issue. Like James Prosek, we must all be wondering ``what can art provide in this time when our earth is indeed looking vulnerable to rapid and regrettable change induced by human influence?''\cite{prosek} If Art is a transcendent mechanism for collapsing the scales of Time and space we need it now more than ever to make sense of this recursively enormous question.

>>>>

The question of what it means to be alive is unavoidable when considering the relationship between movement and what is natural. Movement is a key indicator of life and the ability to distinguish between the living and non-living is a deeply rooted capacity of the human brain. Psychologists and Neurologists have long been aware that ``the distinction between living and non-living things forms a natural place,'' to divide the functionalities and capacities of the mind.\cite{narine} Kinetic sculpture layers these two distinctions -- between the natural and the artificial and the living and non-living.

@@

Finally, it would be an oversight not to mention the hundreds of engineers and makers who devoted much time and thought to the development of OpenSource emergency ventilators during the beginning and worst months of the pandemic.\cite{read} Very few of these devices were ever used on patients, but the collective thought and conversation surrounding the mechanics of breath cannot be discounted as useless, if only as a poignant reminder of the body's mystery and vulnerability.

@@

Storage describes the way air is stored and contained during the act of breath. This is done by the lungs and is only necessary for the transfer of oxygen to the blood. While the lungs are arguably the most critical part of the respiratory system, however they are the most difficult to simulate with accuracy. Interestingly they have relatively little bearing on the sound of breath in this context.

@@

Modulation is the way the sound of breath is shaped as it leaves the body. This is achieved by the bronchial tree, trachea, larynx, pharynx, and mouth.\cite{sarkar} Changes in the shape of these structures, particularly the trachea and the mouth, change the way breath sounds as it leaves the body. The main focus of this project was to reproduce the sound of human breath. After actuation has been achieved, modulation is therefore the most interetsing and relevant part of the respiratory system to investigate.

@@

My two main areas of inquiry, were auditory prototypes -- exploring the effect of disembodied breath noises -- as well as mechanical prototypes -- exploring ways to achieve a physiologically accurate breathing mechanism. In addition to creating a visual tie between the biological nature of the respiratory system, I chose to closely mimic the diaphragm and trachea to prevent the noise of the mechanical components from overwhelming the noise of the breath produced. This approach relies on the material properties of latex and vinyl -- elasticity and rigidity -- to be functional.

@@

To test the effect of disembodied breath on strangers, a small speaker was placed inside of two five gallon plastic buckets. An old cellphone was pre-loaded with a three minute sound effect track of different types of breathing noises. These cycled through breath which sounded like someone sleeping to someone wheezing from exertion. In one bucket, a hole was drilled and a large diameter latex tube was inserted. This provided a way to direct the sound emanating from the speaker and obfuscated the function of the assembly. The assembly was then placed near a crosswalk with a high level of foot traffic around 8 PM ET on the day of Halloween. This location was chosen specifically because it was easy to observe discreetly and had a wide variety of pedestrians.

@@

The objective of this second category of trials was to investigate ways of generating breathing noises mechanically. Using the systems breakdown discussed above, I attempted to create mechanisms for both modulation and actuation of breath. I first approached the diaphragm because actuation occurs before modulation. Using the five gallon buckets, I cut out the center of the lid of one bucket and used it to secure a large piece of 25 mil natural latex over the opening. I then inserted the same large piece of latex tubing into a hole at the other end of the bucket. By depressing the latex, air was forced out of the tube. Pinching the tube at different heights and changing the cadence of the diaphragm movement created a pretty wide range of breath-like noises.

@@

After measuring the force required to depress the lever fully -- 100 Newtons -- I determined that creating an electronic system to activate the diaphragm would be costly and complicated given the scope of time. A few attempts at creating mechanical advantage for the lever were attempted, but none had successful results. It was very difficult to generate the required amount of force without putting a significant amount of weight on the lever itself. I ultimately decided that, due to the expense of high-powered and quiet motors, it made more sense to solve this problem after the core interaction of the piece had been refined.

@@

I embedded four sections of this tubing between two sheets of fourteen mil natural latex. I then bent this into a curved tube-like shape and enclosed the open end of the C. Once this basic shape was complete I threaded each piece of tubing with monofilament and anchored the assembly between two dowels. By pulling on the monofilament is was possible to contract each ring within the tube individually. This result was highly successful and it was possible to contract the aperture to varying degrees.

@@

While this trachea meachanism was not integrated into the modulation prototype, it was by far the most naturalistic movement produced during the investigation. In comparison to the latex tubing used to mimic the trachea earlier, this mechanical latex assembly has much thinner walls. Overall, I believe this will produce more accurate breath noises when combined with the actuation mechanism

@@

I first created a digital breath generator in order to create a more nuanced interaction between the viewer and the breathing noises. This was done using the javascript library P5.js and allowed me to change the cadence, pitch, and volume of the breathing noise. A simple infra-red distance sensor was then connected to the generator. As the distance between the viewer and the work decreased the breathing increased in intensity. The breath noise was amplified using a class D audio amplifier and played through speakers mounted on the front of the piece.

@@

The mechanical form of the work stayed the same in this trial, however I removed the five gallon bucket in favor of a clear vessel which allowed the viewer to see the diaphragm moving. I also added an abstract bronchial tree inside the vessel to suggest lungs. In combination with the added electronics this had a more polished effect.

@@

My initial research into the effects of disembodied breath revealed that the intensity and volume of the breathing noise has a large impact on the comfort level of the viewer. It was clear from the repetition of these trials in different settings that volume played a major role in the level of discomfort. Loud breathing noises inherently felt closer and more intimate than quieter noises. Cadence and pitch of breathing noises also impacted the emotion of the observer. A fast cadence of slightly higher pitch at higher volume was more noticeable and produced a higher degree of discomfort than a slower, softer breathing noise.

In trials which involved playing ``sick'' breathing, viewer response was much more intense. Results of these auditory trials were interesting because they illustrated just how sensitive the group had become to the noise of respiratory distress. While direct comparison is not possible, I wonder if these experiments would have had the same effect before the COVID-19 pandemic.

Understanding the different ways viewers reacted to breathing noises was very helpful as I began to develop the digital breath generator. The ability to dynamically change the intensity and volume of breathing noises based on user proximity created a much more intuitive interaction between the viewer and the breathing noise. In user interviews I found that it was more impactful to have the breath noise increas in intensity as the user approached -- the change in intensity of breathing piqued viewers interest, causing them to approach the work. When they were very close, the breathing reached its highest level of intensity and many viewers noted that they felt somewhat uncomfortable to be in such close proximity to the piece. In comparison, some viewers mentioned that at far away distances -- producing slow and relaxed breathing noises -- they found their own breath falling into a relaxing rhythm and were enjoyed the meditative quality produced. Overall, the auditory interaction consistently caused viewers to consider their own breath.

@@

By replacing the lever mechanism with a crank shaft -- a technique for translating circular motion into linear motion -- I hope to have fine control over the movement of the diaphragm and thereby the actuation of air through the work. This will then allow for further development and testing of the mechanical trachea. The modulation of breath does not end with the trachea however, and I hope to continue exploring the ways in which this sound can be manipulated. Perhaps the development of a mechanical pharynx, or larynx would follow.

@@

The primary objective of this work was to create a biomimetic kinetic sculpture which appeared to be alive and breathing, thereby asking what it means for an non-living, artifical construct to breathe in a natural, life-like way. This objective was approached through a combination of auditory and mechanical explorations. These investigations revealed that both the sound and movement of breath can illicit strong emotions.

As an audible indicator of life, I found that breath can be manipulated to indicate anxiety, calm, or signal respiratory distress. The ongoing COVID-19 pandemic has undoubtably made us more attuned to the subtle changes in breath that signal sickness or health, as was clear from my explorations of breathing noises that included coughing an wheezing. Yet, only a small fraction of the range of emotion that can be conveyed by breath was explored. I hope to extend this work to explore the narrative interactions that can be created using breath.

Moving beyond the sound of breath, this work revealed that the movement of breath is also an intuitive signifier of life. The mechanical goal was to isolate the components of the respiratory system which form the noise created by breathing and highlight their movement. The intensity of viewers' reactions to the mechanical system developed confirmed that movement is just as powerful as, if not more powerful than, sound in indicating that something is alive. Refining the overarching context and quality of the movement will create the inherently kinetic connection between what is natural and what is alive. The medium of kinetic sculpture is uniquely situated to draw this connection, narrowing our focus from the incredible biological complexity of the human body, we can view the natural qualities of breath which indicate life through the most artificial and contrived of lenses.

While the simple and natural act of human breath may be just that, the external acts of human existence which follow breath quickly leave behind the common ground of what is natural. Each human breath on this planet has moved us one step further into the Anthropocene and closer, some would argue to the end of nature.\cite{smith} To conceptualize the impact of human activity on this planet is practically impossible, yet we find ourselves at a juncture where this definition is at the heart of nearly every social and political issue. Like James Prosek, we must all be wondering ``what can art provide in this time when our earth is indeed looking vulnerable to rapid and regrettable change induced by human influence?''\cite{prosek} If Art is a transcendent mechanism for collapsing the scales of Time and space we need it now more than ever to make sense of this recursively enormous question.

% In addition to my intent of using breath as an indicator of life to explore the boundary between the natural and artificial...

% The ongoing COVID-19 pandemic has certainly made us more attuned to the subtle changes in breath that signal respiratory distress.
